Using functor categories to generate intermediate code with Agda

### 2025-05-19 content.tex
Minitoc is back! Hooray!

    \begin{tabular}{ll}
        \textbf{Candidate Number:} & 2330G \\
        \textbf{Title of Project:} & Using functor categories to generate intermediate code with Agda \\
        \textbf{Examination} & Computer Science Tripos - Part II - 2025 \\
        \textbf{Word-count:} & [wordcount] \footnotemark \\
        \textbf{Code line count:} & [linecount] \footnotemark \\
        \textbf{Project Originator:} & Yulong Huang \\
        \textbf{Project Supervisor:} & Yulong Huang and Yufeng Li \\
    \end{tabular}

@@

@@

    \vspace{2em}
    \textbf{\Large Special Difficulties}
    \vspace{1em}
    \begingroup
        \footnotetext[1]{This word-count was computed by \texttt{texcount -1 -sum -merge -q dissertation.tex}.}
        \footnotetext[2]{This code line count was computed by \texttt{find . -name "*.agda" | xargs wc -l}.}
    \endgroup
\end{titlepage}

@@

        \subsection{Contexts and variables}
        We define the context as a finite list of types as follows:
        \[\includegraphics[scale=1.2]{/source/contexts.pdf}\]
        When we are looking for a variable, \gnt{Zero} case corresponds to the situation where the variable is the head of the context, and the \gnt{Suc} case corresponds to the situation where the variable is in the tail of the context. The number formed by \gnt{Zero} and \gnt{Suc} is the index of the variable in the context, called the \emph{de Bruijn index}~\autocite{de_bruijn}.

@@

    \subsubsection{Properties}
    The properties of stack descriptors can be summarised as follows:
    \begin{table}[H]
        \centering
        \begin{tabular}{|c|c|}
            \hline
            \textbf{Agda term} & \textbf{Mathematical meaning} \\
            \hhline{|=|=|}
            \mbt{≤ₛ-refl} & $\forall sd.\ sd \leq_\mathsf{s} sd$ \\
            \hline
            \mbt{≤ₛ-trans} & $\forall sd, sd', sd''.\ sd \leq_\mathsf{s} sd' \land sd' \leq_\mathsf{s} sd'' \Rightarrow sd \leq_\mathsf{s} sd''$ \\
            \hline
            \mbt{+ₛ→≤ₛ} & $\forall sd, n.\ sd \leq_\mathsf{s} sd +_\mathsf{s} n$ \\
            \hline
            \mbt{sub-sd≤ₛ} & $\forall sd, sd', sd''. sd' \equiv sd'' \land sd \leq_\mathsf{s} sd' \Rightarrow sd \leq_\mathsf{s} sd''$ \\
            \hline
            \mbt{–ₛ≡} & $\forall f, d, d′, n.\ d′ - n \equiv d \Rightarrow \ang{f, d} \equiv \ang{f, d′} -_\mathsf{s} n$ \\
            \hline
        \end{tabular}
        \caption{Properties of order and addition of stack descriptors}
        \label{tab: stack_descriptor_properties}
    \end{table}
    % We can use reflexivity and transitivity of the order $<$ and $\leq$ to prove that the order of stack descriptors $\leq_\mathsf{s}$ is a preorder. The proof in Agda is as follows:
    % \begin{proof}
    %     We need to show that $\leq_\mathsf{s}$ is reflexive and transitive. 
    %     \begin{itemize}
    %         \item 
    %             Reflexivity: $\textit{sd} \leq_\mathsf{s} \textit{sd}$ is trivially true since $f = f$ and $d \leq d$.
    %         \item
    %             Transitivity: if $\ang{f, d} \leq_\mathsf{s} \ang{f', d'}$ and $\ang{f', d'} \leq_\mathsf{s} \ang{f'', d''}$, we do the following case split:
    %             \begin{enumerate}
    %                 \item 
    %                     If $f < f'$, $f' = f''$ and $d \leq d'$, then $f < f''$; 
    %                 \item 
    %                     If $f < f'$ and $f' < f''$, then $f < f''$ by transitivity of $<$;
    %                 \item
    %                     If $f = f'$, $d \leq d'$ and $f' < f''$, then $f < f''$;
    %                 \item
    %                     Otherwise $f = f''$ and $d \leq d''$ by transitivity of $\leq$. \footnote{Note that the properties of $=$ is not required in as in the Agda definition of $\leq_\mathsf{s}$, we directly define $f'$ to be $f$ instead of defining a equivalence relation, thus equivalence can be checked directly via type checking.}
    %             \end{enumerate}
    %     \end{itemize}
    % \end{proof}

    % The proof is defined in Agda as follows:
    % \[\includegraphics[scale=1.2]{/target/stack_descriptor_order_proof.pdf}\]

    % When a natural number is added to a stack descriptor, it is guaranteed that the new stack descriptor is not less than the old stack descriptor. This can be proved directly using the same properties of integer addition. The proof is defined in Agda as follows:
    % \[\includegraphics[scale=1.2]{/target/stack_descriptor_add_proof.pdf}\]

@@

    \section{Customised Library} \label{sec: lib}
    The customised library is a collection of types and functions centred around natural numbers, their operations and properties. Instead of using the standard library~\autocite{agda_std}, I implemented a customised library for the following reasons:
    \begin{itemize}
        \item
            \textbf{Specialised Requirements}: The standard library's definitions (e.g. subtraction as monus, `$\dotdiv$') do not align with my project's need for a rigorously defined subtraction operation. This is specified in~\secref{subsec: subtraction}.
        \item
            \textbf{Verification Clarity}: By defining the basic data types and functions from scratch, I can tailor properties (e.g. the property `$n-[n-m] \equiv m$') to my specific requirements, and avoid unnecessary complexity (e.g. `$\leq$' being also defined as a boolean predicate in the standard library). 
        \item
            \textbf{Minimal Dependencies}: Avoiding the standard library reduces external assumptions, making the project self-contained. 
    \end{itemize}
    All definitions and properties can be found in Appendix~[TBD].

    \subsection{Definitions from standard library}
    The following is a list of definitions (or their variants) from the standard library and PLFA~\cite{plfa} that are redefined in the customised library:
    \begin{itemize}
        \item 
            Natural numbers ($\mb{\bN}$), integers ($\mb{\bZ}$) and addition (\mbt{+}).
        \item
            Equality (\mbt{≡}), congruence ($\mb{\mathsf{cong}}$), substitution ($\mb{\mathsf{sub}}$) and transitivity of equality ($\mb{\mathsf{trans}}$).
        \item
            Order of natural numbers, including \mbt{≤} and \mbt{<}.
        \item
            Properties of preorder and addition (summarised in Table~\ref{tab: properties}).
    \end{itemize}
    \begin{table}[H]
        \centering
        \begin{tabular}{|l|l|}
            \hline
            \textbf{Agda term} & \textbf{Mathematical meaning} \\
            \hhline{|=|=|}
            \mbt{+-identity\textsuperscript{r}} & $\forall n \in \bN.\ n + 0 = n$ \\
            \hline
            \mbt{+-suc\textsuperscript{r}}& $\forall m, n \in \bN.\ m + \textsf{suc } n = \textsf{suc } (m + n)$ \\
            \hline
            \mbt{+-comm\textsuperscript{r}} & $\forall m, n \in \bN.\ m + n \equiv n + m$ \\
            \hline
            \mbt{≤-irrelevant} & For all $m, n \in \bN$, we consider all proofs of $m \leq n$ to be equal. \\
            \hline
            \mbt{≤-refl} & $\forall n \in \bN.\ n \leq n$ \\
            \hline
            \mbt{≤-trans} & $\forall m, n, p \in \bN.\ m \leq n \land n \leq p \Rightarrow m \leq p$ \\
            \hline
            \mbt{n≤suc-n} & $\forall n \in \bN.\ n \leq \textsf{suc } n$ \\
            \hline
            \mbt{m≡n,p≤n}\mb{$\to$}\mbt{p≤m} & $\forall p, m, n \in \bN.\ m \equiv n \land p \leq n \Rightarrow p \leq m$ \\
            \hline
            \mbt{+}\mb{$\to$}\mbt{≤} & $\forall m, n \in \bN.\ m \leq m + n$ \\
            \hline
            \mbt{+}\mb{$\to$}\mb{≤\textsuperscript{r}} & $\forall m, n \in \bN.\ m \leq n + m$ \\
            \hline
            \mbt{<-trans} & $\forall m, n, p \in \bN.\ m < n \land n < p \Rightarrow m < p$ \\
            \hline
            \mbt{<}\mb{$\to$}\mbt{s≤} & $\forall m, n \in \bN.\ m < n \Rightarrow \textsf{suc } m \leq n$ \\
            \hline
            \mbt{<}\mb{$\to$}\mbt{≤} & $\forall m, n \in \bN.\ m < n \Rightarrow m \leq n$ \\
            \hline
            \mbt{≤-compare} & $\forall m, n \in \bN.\ m \leq n \lor n \leq m$ \\
            \hline
        \end{tabular}
        \caption{Properties of order and addition}
        \label{tab: properties}
    \end{table}

@@

    \subsection{Subtraction} \label{subsec: subtraction}
    In the standard library~\autocite{agda_std}, the subtraction of natural numbers is defined as follows:\footnote{The definition in the standard library directly uses symbol $\mathsf{-}$, but here for clarity we use $\mathsf{\dotdiv}$ instead.}
    \[\includegraphics[scale=1.2]{minus.pdf}\]
    Mathematically we have $ n \dotdiv m = \max(0, n - m)$. This definition is valid and ensures that the result is a natural number. However, there are two problems with this definition for our implementation:
    \begin{itemize}
        \item 
            The definition of $\dotdiv$ is not suitable for this project. Input $n$ should be guaranteed not to be less than $m$, and at any point in the program if there is a subtraction where $n < m$, the program should fail to type check instead of making $0$ type check with it.
        \item
            We do not have many numerical properties of natural numbers that are required for the implementation of the compiler. For example, $n \dotdiv m + m = n$ or $n \dotdiv (n \dotdiv m) = m$ are not guaranteed to hold if we allow $n < m$. 
    \end{itemize}
    A definition of subtraction that ensures the result is a natural number and preserves the numerical properties is required for the implementation of the compiler. The key is to have a proof that the first argument is greater than or equal to the second argument. There are two approaches to achieve this: passing proofs explicitly or using \textsf{Fin}, a dependent type that encodes the proof of boundedness.

@@

    \section{Compiler} \label{sec: compiler}
    This section develops the semantic foundation of the compiler, starting with the presheaf semantics of the target language. Building on this framework, I introduce the continuation-passing constructs to model control flow, enabling a simplified denotational semantics of types and contexts. I then derive a functorial mapping of semantics terms and the target terms. This section concludes with the implementation of the compiler from the denotational semantics of terms in the source language.
    \subsection{Presheaf semantics}
    The denotation semantics interprets types as presheaves over a base category $\Upsigma$. 
    \subsubsection{Base category}
    The base category $\Upsigma$ is specified as follows:
    \begin{itemize}
        \item Objects are stack descriptors ($\textit{sd} = \ang{f, d}$);
        \item Morphisms are stack expansions ($\iota : \textit{sd} \to \textit{sd′}$ for $\textit{sd} \leq_\mathsf{s} \textit{sd′}$).
    \end{itemize}
    We define $\mathsf{SD}$ as the set of stack descriptors. Then the base category is the category determined by preorder $\underline{\mathsf{SD}} = \{\mathsf{SD}, \leq_\mathsf{s}\}$ in Example~\ref{ex: category_preorder}.

@@

    \subsection{Continuations}
    In Reynolds' model, we construct $\mathsf{Compl}$ and $\mathsf{Intcompl}$\footnote{In the work of Reynolds, $\mathsf{compl}$ and $\mathsf{intcompl}$ are introduced as new types in the source language, and the definitions here are the denotational semantics of those types, respectively. Since these two introduced types are only used for the denotational semantics, I think directly defining them as denotational semantics constructs is natural and sufficient.} as tools in the denotational semantics to represent command continuations and integer continuations, respectively. We can think of integer continuations as a function that takes an integer and returns a command continuation. Mathematically we have
    \[\mathsf{Compl}\ \textit{sd} : \mathsf{SD} \to O \qquad \mathsf{Intcompl}\ \textit{sd} : \bZ \to (\mathsf{SD} \to O)\]
    where $O$ is an unspecified domain of outputs.

@@

    \subsection{Denotational semantics of types and contexts} \label{subsec: type_interpretation}
    With the idea of continuation, the denotational semantics of types can be defined as follows:
    \begin{itemize}
        \item Command passes on command continuations;
        \item Integer expressions takes an integer continuation and returns a command continuation;
        \item Integer acceptors take a command continuation and returns an integer continuation;
        \item Integer variables can be used both as an integer expression and an integer acceptor, so it has the denotational semantics of both, defined as a product of the two.
    \end{itemize}

    The denotational semantics of types in Agda as follows:
    \[\includegraphics[scale=1.2]{/compiler/type_interpretation.pdf}\]

    The denotational semantics of contexts is essentially a list expressed by the product of the denotational semantics of the types in the context. The denotational semantics of contexts is defined as follows, where we use \mb{∅} to denote the denotational semantics of the empty context:
    \[\includegraphics[scale=1.2]{/compiler/context_interpretation.pdf}\]

    \subsection{Functorial mapping} \label{subsec: functorial_mapping}
    The denotational semantics of terms is defined as a functor. Before we define its object mapping, we need to define the functorial mapping, which acts on the morphisms of the base category $\Upsigma$.

@@

        \subsubsection{Variables}
        We define an auxiliary function that simply gets the variable from the list of variables (the context), and the denotational semantics of variables is given by applying this function:
        \[\includegraphics[scale=1.2]{/compiler/term_var.pdf}\]

        \subsubsection{Subtyping}
        The idea of subtyping is to use a term of type $A$ as a term of type $A'$, where $A$ is a subtype of $A'$. We define an auxiliary function that takes in the subtype relation $A\leq:A'$ and the denotational semantics of $A$, and returns the denotational semantics of $A'$. 
        \begin{itemize}
            \item For \gnt{var-≤:-exp} and \gnt{⟦var-≤:-acc}, sincethe denotational semantics of integer variable is defined as product of the denotational semantics of integer expression and integer acceptor in~\secref{subsec: type_interpretation}, we can simply use the projection to get the denotational semantics of either side.
            \item For \gnt{≤:-refl}, we can simply use the identity function.
            \item For \gnt{≤:-trans}, we can use the composition of the two functions.
            \item For \gnt{≤:-fn}, we can use a contravariant function.
        \end{itemize}
        The denotational semantics of subtyping is given by applying the auxiliary function:
        \[\includegraphics[scale=1.2]{/compiler/term_subtype.pdf}\]

@@

        Please see Appendix~[TBD] for the full implementation of the function \textsf{NewVar}.

        \subsubsection{Integer expression} \label{subsubsec: compiler_intexp}
        The denotational semantics of type $\mathsf{intexp}$ is an exponential from an integer continuation to a command continuation. We need to fill the integer continuation with appropriate $R$ terms to get the command continuation. 
        \begin{itemize}
            \item For integer literals, we can simply fill in the integer literal (as an $\mathsf{R}$ term) to the integer continuation $\beta$.
            \item For addition and negation, we wish to fill the negation or the sum of the given expression(s) to the integer continuation $\beta$. However, this is only possible when the given expression(s) are simple left-hand sides, as the grammar of the target language in Figure~\ref{fig: target_grammar} restricts right-hand sides to contain at most one operator. An auxiliary function \textsf{use-temp} that stores the intermediate result of given expressions is introduced to solve this problem. 
        \end{itemize}
        With \textsf{use-temp}, the denotational semantics of addition and negation is simply wrapping $\beta$ with a lambda function using \textsf{use-temp} to get temporary variable(s) of the given expression(s), and directly fill the negation or the sum of the variable(s) to the integer continuation $\beta$. Note that in the denotational semantics of addition, we need to use the functorial mapping of simple right-hand sides for the first argument, as the stack may grow during the evaluation of the second argument. The denotational semantics of integer expressions is defined as follows:
        \[\includegraphics[scale=1.2]{/compiler/term_intexp.pdf}\]
        The auxiliary function \textsf{use-temp} checks if the given expression is a simple left-hand side,
        \begin{itemize}
            \item If it is, we simply fill the expression to the integer continuation $\beta$.\footnote{The type of $\beta$ here is actually a bit different to integer continuation I defined, as it only accepts simple right-hand sides instead of right-hand sides. Reynolds gave a wrong definition of the type of this term. I have corrected it in my Agda implementation.}
            \item If it is not, we need to use a temporary variable to store the result of the expression, and then fill the temporary variable to the integer continuation $\beta$.
        \end{itemize}
        However, regarding the stack descriptor, according to Reynolds' model, we assume
        \begin{itemize}
            \item \textit{sd} is the stack descriptor before the evaluation of plus or negation, which is the position of the temporary variable in the stack.
            \item \textit{sd′} is the stack descriptor after the evaluation on the given expression and just before the assignment of the temporary variable. \textit{sd} ≤ₛ \textit{sd′} is guaranteed.
            \item \textit{sd″} = \textit{sd} +ₛ 1 is the stack descriptor after the assignment of the temporary variable, as it is just large enough to hold the temporary variable.
        \end{itemize}
        In Reynolds' model, the stack descriptor can be directly adjusted from \textit{sd′} to \textit{sd″}. However, in the Agda implementation, we need to know whether the adjustment is positive or negative and use \textsf{assign-inc} or \textsf{assign-dec} respectively. This is done by using \mbt{≤-compare} in Table~\ref{tab: properties}. We also need to show that the adjustment is valid for each case. Please refer to Appendix~[TBD] for the full implementation of the function \textsf{use-temp}.
        % The auxiliary function \textsf{use-temp} is defined as follows:
        % \[\includegraphics[scale=1.2]{/compiler/assign_use-temp.pdf}\]

@@

    The specifications that include multiple variables, free variables and self-referential assignments ensure that the translation of the Debrujin index of the context works correctly under edge cases.

@@

\chapter{Operational Semantics of the source language} \label{app: operational_semantics}
    \chapterminitoc
    The operational semantics of the source language is defined with renaming, substitution and reduction rules. The implementation is as follows:
    \begin{figure}[H]
        \centering
        \includegraphics[scale=1.2]{/source/operational_semantics_1.pdf}
        \caption{Operational semantics of the source language (Part 1)}
        \label{fig: operational_semantics_1}
    \end{figure}
    \begin{figure}[H]
        \centering
        \includegraphics[scale=1.2]{/source/operational_semantics_2.pdf}
        \caption{Operational semantics of the source language (Part 2)}
        \label{fig: operational_semantics_2}
    \end{figure}

>>>>

    \begin{tabular}{ll}
        \textbf{Candidate Number:} & 2330G \\
        \textbf{Title of Project:} & Using functor categories to generate intermediate code with Agda \\
        \textbf{Examination} & Computer Science Tripos - Part II - 2025 \\
        \textbf{Word-count:} & 9574 \footnotemark \\
        \textbf{Code line count:} & 1727 \footnotemark \\
        \textbf{Project Originator:} & Yulong Huang \\
        \textbf{Project Supervisor:} & Yulong Huang and Yufeng Li \\
    \end{tabular}

@@

    The original aims of the project were to implement a compiler originated from the denotational semantics, following the work of Reynolds~\autocite{Reynolds}. The compiler is a functor from the source language to the target language. Agda is chosen as the implementation language because it is a dependently typed proof assistant, which ensures that the input source program, the compilation process, and the output target program are all well-typed.

@@

    I defined a compiler in Agda for the source language and target language as described in Reynolds' work~\autocite{Reynolds}. The compiler was guaranteed to generate expected code by tests in a separate Agda file. An additional customised library was defined to support the implementation of the compiler. More importantly, I spotted a mistake in Reynolds' definition of the denotational semantics of the source language and provided a corrected version. All success criteria were met, and the project was completed successfully.

    \vspace{2em}
    \textbf{\Large Special Difficulties}
    \vspace{1em}

    None.

    \begingroup
        \footnotetext[1]{This word-count was computed by \texttt{texcount -sum -1 content.tex}.}
        \footnotetext[2]{This code line count was computed by \texttt{find . -name "*.agda" | xargs wc -l}.}
    \endgroup
\end{titlepage}

@@

        \subsection{Contexts and variables}
        I define the context as a finite list of types as follows:
        \[\includegraphics[scale=1.2]{/source/contexts.pdf}\]
        When we are looking for a variable, \gnt{Zero} case corresponds to the situation where the variable is the head of the context, and the \gnt{Suc} case corresponds to the situation where the variable is in the tail of the context. The number formed by \gnt{Zero} and \gnt{Suc} is the index of the variable in the context, called the \emph{de Bruijn index}~\autocite{de_bruijn}.

@@

    \subsubsection{Properties}
    The properties of stack descriptors can be summarised as follows:
    \begin{table}[H]
        \centering
        \begin{tabular}{|c|c|}
            \hline
            \textbf{Agda term} & \textbf{Mathematical meaning} \\
            \hhline{|=|=|}
            \mbt{≤ₛ-refl} & $\forall sd.\ sd \leq_\mathsf{s} sd$ \\
            \hline
            \mbt{≤ₛ-trans} & $\forall sd, sd', sd''.\ sd \leq_\mathsf{s} sd' \land sd' \leq_\mathsf{s} sd'' \Rightarrow sd \leq_\mathsf{s} sd''$ \\
            \hline
            \mbt{+ₛ→≤ₛ} & $\forall sd, n.\ sd \leq_\mathsf{s} sd +_\mathsf{s} n$ \\
            \hline
            \mbt{sub-sd≤ₛ} & $\forall sd, sd', sd''. sd' \equiv sd'' \land sd \leq_\mathsf{s} sd' \Rightarrow sd \leq_\mathsf{s} sd''$ \\
            \hline
            \mbt{–ₛ≡} & $\forall f, d, d′, n.\ d′ - n \equiv d \Rightarrow \ang{f, d} \equiv \ang{f, d′} -_\mathsf{s} n$ \\
            \hline
        \end{tabular}
        \caption{Properties of order and addition of stack descriptors}
        \label{tab: stack_descriptor_properties}
    \end{table}
    The definitions of the above properties can be found in Appendix~\ref{app: target_properties}.

@@

    \section{Customised Library} \label{sec: lib}
    The customised library is a collection of types and functions centred around natural numbers, their operations and properties. Instead of using the standard library~\autocite{agda_std}, I implemented a customised library for the following reasons:
    \begin{itemize}
        \item
            \textbf{Specialised Requirements}: The standard library's definitions (e.g. subtraction as monus, `$\dotdiv$') do not align with my project's need for a rigorously defined subtraction operation. This is specified in~\secref{subsec: subtraction}.
        \item
            \textbf{Verification Clarity}: By defining the basic data types and functions from scratch, I can tailor properties (e.g. the property `$n-[n-m] \equiv m$') to my specific requirements, and avoid unnecessary complexity (e.g. `$\leq$' being also defined as a boolean predicate in the standard library). 
        \item
            \textbf{Minimal Dependencies}: Avoiding the standard library reduces external assumptions, making the project self-contained. 
    \end{itemize}
    All definitions and properties that are listed in this section can be fount in Appendix~\ref{app: lib_properties}.

    \subsection{Definitions from standard library}
    The following is a list of definitions (or their variants) from the standard library and PLFA~\cite{plfa} that are redefined in the customised library:
    \begin{itemize}
        \item 
            Natural numbers ($\mb{\bN}$), integers ($\mb{\bZ}$) and addition (\mbt{+}).
        \item
            Equality (\mbt{≡}), congruence ($\mb{\mathsf{cong}}$), substitution ($\mb{\mathsf{sub}}$) and transitivity of equality ($\mb{\mathsf{trans}}$).
        \item
            Order of natural numbers, including \mbt{≤} and \mbt{<}.
        \item
            Properties of preorder and addition (summarised in Table~\ref{tab: properties}).
    \end{itemize}
    The properties are defined in a minimalistic way and tailored to the requirements of the compiler. For example, \mbt{≤-compare} in Table~\ref{tab: properties} is a property that states that for all $m, n \in \bN$, either $m \leq n$ or $n \leq m$, which is used to case split in the implementation of the compiler. In the standard library, a similar property is defined to split the cases into $m < n$, $m \equiv n$ and $m > n$, which adds unnecessary complexity to the implementation, as all the proof required for compiler is in the form of $m \leq n$.
    \begin{table}[H]
        \centering
        \begin{tabular}{|l|l|}
            \hline
            \textbf{Agda term} & \textbf{Mathematical meaning} \\
            \hhline{|=|=|}
            \mbt{+-identity\textsuperscript{r}} & $\forall n \in \bN.\ n + 0 = n$ \\
            \hline
            \mbt{+-suc\textsuperscript{r}}& $\forall m, n \in \bN.\ m + \textsf{suc } n = \textsf{suc } (m + n)$ \\
            \hline
            \mbt{+-comm\textsuperscript{r}} & $\forall m, n \in \bN.\ m + n \equiv n + m$ \\
            \hline
            \mbt{≤-irrelevant} & For all $m, n \in \bN$, we consider all proofs of $m \leq n$ to be equal. \\
            \hline
            \mbt{≤-refl} & $\forall n \in \bN.\ n \leq n$ \\
            \hline
            \mbt{≤-trans} & $\forall m, n, p \in \bN.\ m \leq n \land n \leq p \Rightarrow m \leq p$ \\
            \hline
            \mbt{n≤suc-n} & $\forall n \in \bN.\ n \leq \textsf{suc } n$ \\
            \hline
            \mbt{m≡n,p≤n}\mb{$\to$}\mbt{p≤m} & $\forall p, m, n \in \bN.\ m \equiv n \land p \leq n \Rightarrow p \leq m$ \\
            \hline
            \mbt{+}\mb{$\to$}\mbt{≤} & $\forall m, n \in \bN.\ m \leq m + n$ \\
            \hline
            \mbt{+}\mb{$\to$}\mb{≤\textsuperscript{r}} & $\forall m, n \in \bN.\ m \leq n + m$ \\
            \hline
            \mbt{<-trans} & $\forall m, n, p \in \bN.\ m < n \land n < p \Rightarrow m < p$ \\
            \hline
            \mbt{<}\mb{$\to$}\mbt{s≤} & $\forall m, n \in \bN.\ m < n \Rightarrow \textsf{suc } m \leq n$ \\
            \hline
            \mbt{<}\mb{$\to$}\mbt{≤} & $\forall m, n \in \bN.\ m < n \Rightarrow m \leq n$ \\
            \hline
            \mbt{≤-compare} & $\forall m, n \in \bN.\ m \leq n \lor n \leq m$ \\
            \hline
        \end{tabular}
        \caption{Properties of order and addition}
        \label{tab: properties}
    \end{table}

@@

    \subsection{Subtraction} \label{subsec: subtraction}
    In the standard library~\autocite{agda_std}, the subtraction of natural numbers is defined as follows:\footnote{The definition in the standard library directly uses symbol $\mathsf{-}$, but here for clarity I use $\mathsf{\dotdiv}$ instead.}
    \[\includegraphics[scale=1.2]{minus.pdf}\]
    Mathematically we have $ n \dotdiv m = \max(0, n - m)$. This definition is valid and ensures that the result is a natural number. However, there are two problems with this definition for our implementation:
    \begin{itemize}
        \item 
            The definition of $\dotdiv$ is not suitable for this project. Input $n$ should be guaranteed not to be less than $m$, and at any point in the program if there is a subtraction where $n < m$, the program should fail to type check instead of making $0$ type check with it.
        \item
            We do not have many numerical properties of natural numbers that are required for the implementation of the compiler. For example, $n \dotdiv m + m = n$ or $n \dotdiv (n \dotdiv m) = m$ are not guaranteed to hold if we allow $n < m$. 
    \end{itemize}
    A definition of subtraction that ensures the result is a natural number and preserves the numerical properties is required for the implementation of the compiler. The key is to have a proof that the first argument is greater than or equal to the second argument. There are two approaches to achieve this: passing proofs explicitly or using \textsf{Fin}, a dependent type that encodes the proof of boundedness.

@@

    \section{Compiler} \label{sec: compiler}
    This section develops the semantic foundation of the compiler, starting with the presheaf semantics of the target language. Building on this framework, I introduce the continuation-passing constructs to model control flow, enabling a simplified denotational semantics of types and contexts. I then derive a functorial mapping of semantics terms and the target terms. This section concludes with the implementation of the compiler from the denotational semantics of terms in the source language.
    \subsection{Presheaf semantics}
    The denotation semantics interprets types as presheaves over a base category $\Upsigma$. 
    \subsubsection{Base category}
    The base category $\Upsigma$ is specified as follows:
    \begin{itemize}
        \item Objects are stack descriptors ($\textit{sd} = \ang{f, d}$);
        \item Morphisms are stack expansions ($\iota : \textit{sd} \to \textit{sd′}$ for $\textit{sd} \leq_\mathsf{s} \textit{sd′}$).
    \end{itemize}
    I define $\mathsf{SD}$ as the set of stack descriptors. Then the base category is the category determined by preorder $\underline{\mathsf{SD}} = \{\mathsf{SD}, \leq_\mathsf{s}\}$ in Example~\ref{ex: category_preorder}.

@@

    \subsection{Continuations}
    In Reynolds' model, $\mathsf{Compl}$ and $\mathsf{Intcompl}$\footnote{In the work of Reynolds, $\mathsf{compl}$ and $\mathsf{intcompl}$ are introduced as new types in the source language, and the definitions here are the denotational semantics of those types, respectively. Since these two introduced types are only used for the denotational semantics, I think directly defining them as denotational semantics constructs is natural and sufficient.} are constructed as tools in the denotational semantics to represent command continuations and integer continuations, respectively. We can think of integer continuations as a function that takes an integer and returns a command continuation. Mathematically we have
    \[\mathsf{Compl}\ \textit{sd} : \mathsf{SD} \to O \qquad \mathsf{Intcompl}\ \textit{sd} : \bZ \to (\mathsf{SD} \to O)\]
    where $O$ is an unspecified domain of outputs.

@@

    \subsection{Denotational semantics of types and contexts} \label{subsec: type_interpretation}
    With the idea of continuation, the denotational semantics of types can be defined as follows:
    \begin{itemize}
        \item Commands pass on command continuations;
        \item Integer expressions take integer continuations and return command continuations;
        \item Integer acceptors take command continuations and return integer continuations;
        \item Integer variables can be used either as integer expressions or integer acceptors, so they have the denotational semantics of both, defined as a product of the two.
    \end{itemize}

    The denotational semantics of types in Agda is as follows:
    \[\includegraphics[scale=1.2]{/compiler/type_interpretation.pdf}\]

    The denotational semantics of contexts is essentially a list expressed by the product of the denotational semantics of the types in the context. The denotational semantics of contexts is defined as follows, where I use \mb{∅} to denote the denotational semantics of the empty context:
    \[\includegraphics[scale=1.2]{/compiler/context_interpretation.pdf}\]

    \subsection{Functorial mapping} \label{subsec: functorial_mapping}
    The denotational semantics of terms is defined as a functor. Before defining its object mapping, we need to define the functorial mapping, which acts on the morphisms of the base category $\Upsigma$.

@@

        \subsubsection{Variables}
        I define an auxiliary function that simply gets the variable from the list of variables (the context), and the denotational semantics of variables is given by applying this function:
        \[\includegraphics[scale=1.2]{/compiler/term_var.pdf}\]

        \subsubsection{Subtyping}
        The idea of subtyping is to use a term of type $A$ as a term of type $A'$, where $A$ is a subtype of $A'$. I define an auxiliary function that takes in the subtype relation $A\leq:A'$ and the denotational semantics of $A$, and returns the denotational semantics of $A'$. 
        \begin{itemize}
            \item For \gnt{var-≤:-exp} and \gnt{⟦var-≤:-acc}, since the denotational semantics of integer variable is defined as product of the denotational semantics of integer expression and integer acceptor in~\secref{subsec: type_interpretation}, we can simply use the projection to get the denotational semantics of either side.
            \item For \gnt{≤:-refl}, we can simply use the identity function.
            \item For \gnt{≤:-trans}, we can use the composition of the two functions.
            \item For \gnt{≤:-fn}, we can use a contravariant function.
        \end{itemize}
        The denotational semantics of subtyping is given by applying the auxiliary function:
        \[\includegraphics[scale=1.2]{/compiler/term_subtype.pdf}\]

@@

        Please see Appendix~\ref{app: compiler_details} for the full implementation of the function \textsf{NewVar}.

        \subsubsection{Integer expression} \label{subsubsec: compiler_intexp}
        The denotational semantics of type $\mathsf{intexp}$ is an exponential from an integer continuation to a command continuation. We need to fill the integer continuation with appropriate $R$ terms to get the command continuation. 
        \begin{itemize}
            \item For integer literals, we can simply fill in the integer literal (as an $\mathsf{R}$ term) to the integer continuation $\beta$.
            \item For addition and negation, we wish to fill the negation or the sum of the given expression(s) to the integer continuation $\beta$. However, this is only possible when the given expression(s) are simple left-hand sides, as the grammar of the target language in Figure~\ref{fig: target_grammar} restricts right-hand sides to contain at most one operator. An auxiliary function \textsf{use-temp} that stores the intermediate result of given expressions is introduced to solve this problem. 
        \end{itemize}
        With \textsf{use-temp}, the denotational semantics of addition and negation is simply wrapping $\beta$ with a lambda function using \textsf{use-temp} to get temporary variable(s) of the given expression(s), and directly fill the negation or the sum of the variable(s) to the integer continuation $\beta$. Note that in the denotational semantics of addition, we need to use the functorial mapping of simple right-hand sides for the first argument, as the stack may grow during the evaluation of the second argument. The denotational semantics of integer expressions is defined as follows:
        \[\includegraphics[scale=1.2]{/compiler/term_intexp.pdf}\]
        The auxiliary function \textsf{use-temp} checks if the given expression is a simple left-hand side,
        \begin{itemize}
            \item If it is, we simply fill the expression to the integer continuation $\beta$.\footnote{The type of $\beta$ here is actually a bit different to integer continuation I defined, as it only accepts simple right-hand sides instead of right-hand sides. Reynolds gave a wrong definition of the type of this term. I have corrected it in my Agda implementation.}
            \item If it is not, we need to use a temporary variable to store the result of the expression, and then fill the temporary variable to the integer continuation $\beta$.
        \end{itemize}
        However, regarding the stack descriptor, according to Reynolds' model, we assume
        \begin{itemize}
            \item \textit{sd} is the stack descriptor before the evaluation of plus or negation, which is the position of the temporary variable in the stack.
            \item \textit{sd′} is the stack descriptor after the evaluation on the given expression and just before the assignment of the temporary variable. \textit{sd} ≤ₛ \textit{sd′} is guaranteed.
            \item \textit{sd″} = \textit{sd} +ₛ 1 is the stack descriptor after the assignment of the temporary variable, as it is just large enough to hold the temporary variable.
        \end{itemize}
        In Reynolds' model, the stack descriptor can be directly adjusted from \textit{sd′} to \textit{sd″}. However, in the Agda implementation, we need to know whether the adjustment is positive or negative and use \textsf{assign-inc} or \textsf{assign-dec} respectively. This is done by using \mbt{≤-compare} in Table~\ref{tab: properties}. We also need to show that the adjustment is valid for each case. Please refer to Appendix~\ref{app: compiler_details} for the full implementation of the function \textsf{use-temp}.
        % The auxiliary function \textsf{use-temp} is defined as follows:
        % \[\includegraphics[scale=1.2]{/compiler/assign_use-temp.pdf}\]

@@

    The specifications that include multiple variables, free variables and self-referential assignments ensure that the translation of the de Bruijn index of the context works correctly under edge cases.

@@

\chapter{Operational Semantics of the source language} \label{app: operational_semantics}
    The operational semantics of the source language is defined with renaming, substitution and reduction rules. The implementation is as follows:
    \begin{figure}[H]
        \centering
        \includegraphics[scale=1.2]{/appendix/source_operational_semantics_1.pdf}
        \caption{Operational semantics of the source language (Part 1)}
        \label{fig: operational_semantics_1}
    \end{figure}
    \begin{figure}[H]
        \centering
        \includegraphics[scale=1.2]{/appendix/source_operational_semantics_2.pdf}
        \caption{Operational semantics of the source language (Part 2)}
        \label{fig: operational_semantics_2}
    \end{figure}
